Higher-Order unification for free!

### 2024-05-13 paper/main.tex
fm -> to, fapp -> o-app, flam -> o-app
typos

In order to implement such a search in Elpi, we shall describe the encoding
of Coq terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher-Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
terms, called \elpiIn{tm}, that includes (among others) the following
constructors:

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots n,$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
$$
\end{proposition}
\noindent
In particular, this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related, and in turn, this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

\begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
In the context of \hrun, if 
$\rho_{p-1} \foUnifPb_{p} \subseteq \llambda$ 
(even if $\;\foUnifPb_{p} \not\subseteq \llambda$)
then
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p} \Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p}
$$
\end{proposition}

@@

\begin{definition}[\maybeeta]\label{def:maybeeta}
  \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho \lambda x.s$
  is an eta expansion. % without the lambda, 5.11 is not true, that
  % is we assume variables are in W
\end{definition}

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda x.\lambda y.f \appsep y \appsep x\}$
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta-long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\noindent
A solution to a \Fo{} variable is a plain term, that is, \elpiIn{fsubst}
is an abbreviation for \elpiIn{mem fm}.
%We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
%the one of \Ho.

@@

%Applying the substitution corresponds to dereferencing a term with respect to
%the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{o-app[o-app L1|L2]}>> are replaced by
% <<\elpiIn{o-app L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{o-app} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth examining the code of \elpiIn{deref}, which
applies the substitution to a \Ho{} term. Notice how assignments are moved
to the current scope, i.e. the \elpiIn{abs}-bound variables are renamed
with the names in the scope of the unification variable occurrence.

@@

The simple compiler described in this section serves as a basis for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
using %the routine
\ref{clause:malloc}.

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems), which plays no role in this section
but plays a major role in \cref{sec:llam,sec:eta,sec:beta}.
With respect to \cref{sec:problem-statement}, the signature also allows
for updates to the substitution.

@@

\noindent
The code above uses that possibility
in order to allocate space for the variables, i.e. it sets their memory
address to \elpiIn{none} (a detail not worth mentioning in the
previous sections).

@@

A step in \Ho consists of unifying two terms and reconsidering all
links for progress. If either of these tasks fails, we consider the entire step to
fail. It is at this granularity that we can relate steps in the
two languages.

@@

\subsubsection{Progress}
In the base compilation scheme, \elpiIn{progress1} is the identity function
on both the links and the substitution, so the fixpoint trivially terminates.
Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
and explain why the don't hinder termination.
% For brevity we omit the code
% that applies the substitution \elpiIn{S1} to all terms in \linkStore.

\subsubsection{Scope check}\label{sec:sc}
The predicate \elpiIn{scope-check} 
replaces each
link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} with
a link
\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ is obtained after
executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$ using a fresh $Y$.
This unification
ensures that $t'$ cannot contain variables in $\Gamma / \{x_1\ldots x_n\}$,
and if it fails then \progress hence \hstep fails.
In turn this grants fidelity in the case where the \lhs of an \linketa
is pruned of a name that occurs (rigidly) in $t$:
links represent suspended unification problems
\emph{that may succeed}.

@@

First and foremost, problematic terms stored in
\linkStore have to be moved back into the game:
a suspended link must be turned into a valid assignment.
This operation is possible thanks to \fullref{inv:linklhs},
which ensures that no link causes an occur-check~(\ref{sec:oc})
and the fact that \linkStore is duplicate-free (see \cref{sec:invariant1}).

@@

\noindent
So far the compiler is very basic. It does not really
enforce that the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first-order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will ensure \fullref{prop:w-enforcing}
and
the proof will be refined to cover for the new cases.

@@

In the following sections we use the following notation for input and output
of the compiler. \foUnifPb represents the input unification problems in \Fo{}
while \hoUnifPb their corresponding compiled version with memory mapping
\mapStore and links \linkStore. For example:
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
  {{{\eta,\Gamma,a,b}}}

@@

  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms, there is one more interesting case,
  namely \elpiIn{o-app[o-uva X|W] ~\Uo~s} when \elpiIn{W}
  are distinct names compiled to $\vec{w}$. In this case
  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$, which in turn
   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
  \end{proof}

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
% \begin{proof}[Proof sketch]
%   problematic terms are mapped to uva by comp,
%   the problematic o-app node is gone.
%   if F x = t then  F = lam x.t[]
%   \todo{finish/maybe W enforcement is not needed here}
% \end{proof}
\noindent
In other word the compiler eliminates all subterms that are in \maybebeta.

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$.
\label{th:fidelity-beta}
\end{theorem}
\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in $\wellb_\beta$, as well as in \wellb by
  \cref{thm:fidfow}.
\end{proof}

@@

\noindent
The reason is that \elpiIn{lam x\ uva A [x]} and
\elpiIn{con"f"} start with different, rigid, term constructors.

@@

\begin{invariant}[\linketa \rhs]
  The \rhs of any \linketa %in \linkStore 
  has the shape $\lambda x.t$
  and $t$ is \wellb. 
  %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
  \label{inv:link-eta-right}
\end{invariant}

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w \in W$, such that $w$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case, the term $t$ under the spine of binders
$x_1\ldots x_n$ can either be $x$ applied to terms that
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to the application $x \appsep x_1 \ldots x_n$ (case 3).
\end{proof}

@@

In other words, $x$ \occursrigid in $t$ if it occurs in $t$
outside of the scope of a unification variable $X$; otherwise, an instantiation
of $X$ can make $x$ disappears from $t$.
Moreover, note that $\eta$-contracting $t$ cannot make $x$ disappear, since
$x$ is not a locally bound variable inside $t$.

@@

\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or a name applied to the arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $m - n < i \leq m$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term
  and
  $t \in \maybeeta$ then \testmaybeeta{} $t$ holds.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\linketa{}s are meant to delay the unification of ``problematic'' terms until
we know for sure if their head lambdas can be $\eta$-contracted or not.

\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{t} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $t$, that is we run
    $\lambda x.X \appsep x\Ue{} t$
    \item if $X = \lambda x.s$ we run $X \Ue{} t$.
\end{enumerate}
\end{definition}

@@

% \noteD[inline]{Below the proof of \cref{prop:w-preservation}, ho usato 3 lemmi ausiliari,
% forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaleft is between
  terms in $\wellb$  
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be a substitution such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
  By \cref{inv:link-eta-right}, we have $\wellb(t)$.
  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$ and both terms are in \wellb.
  Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
  are both in \wellb.
\end{proof}

@@

\begin{proof}
  Given two \linketa{}s \linketaM{\Gamma_1}{X}{\lambda x.t} and
  \linketaM{\Gamma_2}{Y}{\lambda x.t'} 
  the unification is performed between $t$ and $t'$.
  By \cref{lem:w-enforcement-maybeeta} both terms are
  \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch] %\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo{} unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compiled to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to rule \ref{clause:lam-lam}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
          since rule \ref{clause:lam-lam} or rule \ref{clause:eta1} try to unify under it, eventually finding two
          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

>>>>

In order to implement such a search in Elpi, we shall describe the encoding
of Coq terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher-Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
Coq terms, called \elpiIn{tm}, that includes (among others) the following
constructors:

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of  \frun and \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots n,$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
$$
\end{proposition}
\noindent
In particular, this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related, and in turn, this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

\begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
In the context of \frun and \hrun, if 
$\rho_{p-1} \foUnifPb_{p} \subseteq \llambda$ 
(even if $\;\foUnifPb_{p} \not\subseteq \llambda$)
then
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p} \Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p}
$$
\end{proposition}

@@

\begin{definition}[\maybeeta]\label{def:maybeeta}
  \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho (\lambda x.s)$
  is an eta expansion. % without the lambda, 5.11 is not true, that
  % is we assume variables are in W
\end{definition}

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda x.\lambda y.f \appsep y \appsep x\}$
makes $\rho s \Eo{} \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta-long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\noindent
A solution to a \Fo{} variable is, instead, a plain term, that is, \elpiIn{fsubst}
is an abbreviation for \elpiIn{mem to}.
%We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
%the one of \Ho.

@@

%Applying the substitution corresponds to dereferencing a term with respect to
%the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{o-app[o-app L1|L2]}>> are replaced by
% <<\elpiIn{o-app L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{o-app} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth examining the code of \elpiIn{deref}, that
applies the substitution to a \Ho{} term. Notice how assignments are moved
to the current scope, i.e. the \elpiIn{abs}-bound variables are renamed
with the names in the scope of the unification variable occurrence.

@@

The simple compiler described in this section serves as a base for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones.
% using the routine \ref{clause:malloc}.

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems), which plays no role in this section
but plays a major role in \cref{sec:llam,sec:eta,sec:beta}.

@@

\noindent
With respect to \cref{sec:problem-statement}, the signature also allows
for updates to the substitution.
The code above uses that possibility
in order to allocate space for the variables, i.e. it sets their memory
address to \elpiIn{none} (a detail not worth mentioning in the
previous sections).

@@

A step in \Ho consists in unifying two terms and reconsidering all
links for progress. If either of these tasks fails, we consider the entire step to
fail. It is at this granularity that we can relate steps in the
two languages.

@@

\subsubsection{Progress}
In the base compilation scheme, \elpiIn{progress1} is the identity function
on both the links and the substitution, so the fixpoint trivially terminates.
Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
and explain why the do not hinder termination.
% For brevity we omit the code
% that applies the substitution \elpiIn{S1} to all terms in \linkStore.

\subsubsection{Scope check}\label{sec:sc}
The predicate \elpiIn{scope-check} 
replaces each
link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} with
a link
\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
The term $t' = \lambda x.Y_{{x_1\ldots x_n}~x}$ is obtained after
executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$ using a fresh $Y$.
This unification
ensures that $t'$ cannot contain variables in $\Gamma / \{x_1\ldots x_n\}$,
and if it fails then \progress hence \hstep fails.
In turn this grants fidelity in the case where the \lhs of an \linketa
is pruned of a name that occurs (rigidly) in $t$:
links represent suspended unification problems
\emph{that may succeed}.

@@

First and foremost, problematic terms stored in
\linkStore have to be moved back into the game:
a suspended link must be turned into a valid assignment.
This operation is possible thanks to \fullref{inv:linklhs},
and that no link causes an occur-check~(\ref{sec:oc})
and the fact that \linkStore is duplicate-free (see \cref{sec:invariant1}).

@@

\noindent
So far the compiler is very basic. It does not really
enforce that the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first-order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will ensure \fullref{prop:w-enforcing},
the proof will be refined to cover for the new cases.

@@

In the following sections we use the following notation for input and output
of the compiler. \foUnifPb represents the input unification problems in \Fo{}
while \hoUnifPb their corresponding compiled version with memory mapping
\mapStore and links \linkStore. For example:
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{n}}},{X_2,A_2,{{m}}}}}
  {{{\eta,\Gamma,a,b}}}

@@

  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms, there is one more interesting case,
  namely \elpiIn{o-app[o-uva X|W] ~\Uo~s} when \elpiIn{W}
  are distinct names compiled to $\vec{w}$. In this case
  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$, which in turn
   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1} we have
   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
  \end{proof}

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
% \begin{proof}[Proof sketch]
%   problematic terms are mapped to uva by comp,
%   the problematic o-app node is gone.
%   if F x = t then  F = lam x.t[]
%   \todo{finish/maybe W enforcement is not needed here}
% \end{proof}
\noindent
In other words the compiler eliminates all subterms that are in \maybebeta.

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$.
\label{th:fidelity-beta}
\end{theorem}
\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta},
  \Ue is as powerful as \Uo in $\wellb_\beta$, as well as in \wellb by
  \cref{thm:fidfow}.
\end{proof}

@@

\noindent
The reason is that <<\elpiIn{lam x\ uva A [x]}>> and
<<\elpiIn{con"f"}>> start with different term constructors.

@@

\begin{invariant}[\linketa \rhs]
  The \rhs of any \linketa %in \linkStore 
  has the shape $\lambda x.t$
  and $t$ is in \wellb. 
  %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
  \label{inv:link-eta-right}
\end{invariant}

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w \in W$, such that $w$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form the only
rule that can play a role is \ref{clause:eta1}, and note that
that rule uses \elpiIn{beta} to add an argument to an
application or to the scope of a variable.
If the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case, the term $t$ under the spine of binders
$x_1\ldots x_n$ can either be $x$ applied to terms that
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to the application $x \appsep x_1 \ldots x_n$ (case 3).
\end{proof}

@@

In other words, $x$ \occursrigid in $t$ if it occurs in $t$
outside of the scope of a unification variable $X$; otherwise, an instantiation
of $X$ can make $x$ disappears from $t$.
Note that $\eta$-contracting $t$ cannot make $x$ disappear, since
$x$ is not a locally bound variable inside $t$.

@@

\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant $c$ or a name $y$ applied to the arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $m - n < i \leq m$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n} ~ y$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term
  and
  $t \in \maybeeta$ then \testmaybeeta{} $t$ holds.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\linketa{}s are meant to delay the unification of ``problematic'' terms until
we know for sure if their head lambdas can be $\eta$-contracted.

\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{t} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$. There are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $t$, that is we run
    $\lambda x.X \appsep x\Ue{} t$
    \item if $X = \lambda x.s$ we run $X \Ue{} t$.
\end{enumerate}
\end{definition}

@@

% \noteD[inline]{Below the proof of \cref{prop:w-preservation}, ho usato 3 lemmi ausiliari,
% forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaleft is between
  terms in \wellb.
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be a substitution such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
  By \cref{inv:link-eta-right}, we have $\wellb(t)$.
  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$ and both terms are in \wellb.
  Otherwise, $\lambda x. X\appsep x$ is unified with $\lambda x. t$, and $X\appsep x$ and $t$
  are both in \wellb.
\end{proof}

@@

\begin{proof}
  Given two \linketa{}s \linketaM{\Gamma_1}{X}{\lambda x.t} and
  \linketaM{\Gamma_2}{Y}{\lambda x.t'} 
  the unification is performed between $t$ and $t'$.
  By \cref{lem:w-enforcement-maybeeta} both terms are in
  \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is between
  terms in \wellb.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  %\footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch] %\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo{} unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the lambdas are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.

    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compiled to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft.
          If $s_1$ and $s_2$ are equal thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they are equal thanks to rule \ref{clause:lam-lam}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.
          Alternatively, if $s_1$ are different $s_2$
          it cannot be because of the $\lambda$ constructor in the head of $t_1$,
          since rule \ref{clause:lam-lam} or rule \ref{clause:eta1} move under it, eventually finding two
          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.
